\texttt{Taxonomic supertree construction with}\texttt{\emph{ Incertae sedis}}\texttt{ taxa}

### 2018-04-19 doc/incertae_sedis.tex
minor

\documentclass[english]{article}
\usepackage[T1]{fontenc}
\usepackage[latin9]{inputenc}
\usepackage{geometry}
\geometry{verbose,tmargin=3cm,bmargin=3cm,lmargin=3cm,rmargin=3cm}
\setlength{\parskip}{6pt}
\setlength{\parindent}{0pt}
\usepackage{verbatim}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{graphicx}
\usepackage[authoryear]{natbib}

@@

\usepackage{babel}
\usepackage{xspace}
\usepackage{color}
\begin{document}

@@

Because of the large number (over 2.6 million) of tips in the full
tree, the supertree method of \citet{redelings2017supertree} uses an
approximation approach that relies on decomposition of the full
problem into subproblems. The decomposition partitioning the list of
taxonomy splits $S(\taxonomy_{p})$ into two lists:
$\mathcal{Z}(\taxonomy_{p}),$the list of splits from the taxonomy tree
which are compatible with every rooted split in the phylogenetic
inputs, and $\mathcal{Y}(\taxonomy_{p})$, the list of taxonomic splits
that conflict with at least one phylogenetic input split. Conceptually
the approximate algorithm works by greedily adding all compatible
splits in altered ranked list of splits:
$\mathcal{S}^{\prime}=\mathcal{Z}(\taxonomy_{p})+S(T_{1})+S(T_{2})+\l
dots+S(T_{n})+\mathcal{Y}(\taxonomy_{p})$.

@@

We define rooted splits for each edge by noting that removing the edge
would produce two trees. The leaf label sets of these trees the
partition the full leaf set $\mathcal{L}$ into two groups. The include
set $\mathcal{I}(e)$ which does not contain the root, and the exclude
set $\mathcal{E}(e)$ which does contain the root. Such a split may be
written \[ \mathcal{I}(e)|\bullet\mathcal{E}(e). \] where the
$\bullet$ indicates the root. The include set is the set of leaves
that are descendants of the daughter node of the edge. The exclude set
is the set of leaves that are not descendants of that node. If no taxa
are \emph{incertae sedis}, then the exclude set for a node is just the
total tip set minus the include set for the node: \begin{align*}
\mathcal{E}(e) & =\mathcal{L}-\mathcal{I}(e). \end{align*}

@@

We seek a semantics for supertrees with \emph{incertae sedis }taxa
that satisfy the following properties:
\begin{enumerate}
    \item An
\emph{incertae sedis} node may intrude into its siblings and their
descendants.
\item Two \emph{incertae sedis} siblings may not be
interdigitated.
\item Any placement $T\to T^{*}$ of a tree $T$ with
\emph{incertae sedis} taxa is also accessible from a tree $T^{\prime}$
that is created from $T$ by taking a taxon, attaching it to an
ancestor, and marking it \emph{incertae sedis}.
\begin{comment} MTH:
I don't understand this \end{comment} \item The semantics of an
\emph{incertae sedis} node does not depend on assigned ranks, but only
on the tree.
\item Semantics is based on deriving a split for each
branch of the tree.
\end{enumerate}
Here we focus on a split-based
semantics instead of a semantics based on apply of a series of edit
operations to the tree. The split-based semantics allows us to seek
trees that are simultaneously consistent with a set of split-
constraints. The use of a split-based semantics does impose some
limitations. These come primarily because in the edit-based semantics,
after we place an \emph{incertae sedis} taxon, it is no loner
\emph{incertae sedis}, whereas with the split-based semantics taxa are
either always \emph{incertae sedis} or never \emph{incertae sedis}.

@@

In order to apply supertree methods to rooted trees with
\emph{incertae sedis} taxa, we first propose a semantics of
\emph{incertae sedis }in terms of rooted splits. We define rooted
splits by noting that each edge of a tree divides the tip taxa
$\mathcal{L}$ into two groups: the include set $\mathcal{I}(e)$ which
does not contain the root, and the exclude set $\mathcal{E}(e)$ which
does contain the root. Such a split may be written \[
\mathcal{I}(e)|\bullet\mathcal{E}(e). \] where the $\bullet$ indicates
the root. If no taxa are \emph{incertae sedis}, then the exclude set
for a node is just the total tip set minus the include set for the
node: \begin{align*} \mathcal{E}(n) & =\mathcal{L}-\mathcal{I}(n).
\end{align*}

@@

The reduced exclude set $\mathcal{E}(n)$ for node $n$ should therefore
not contain the descendants of its siblings which are marked
\emph{incertae }sedis\emph{,} but should contain the descendants of
its other siblings\emph{. }We additionally stipulate that it is not
possible for \emph{incertae sedis} taxa to intrude into their
\emph{incertae sedis} siblings. (This is not the only way to construct
a split-based semantics for \emph{incertae sedis}, but it avoids
certain complications, as noted in the discussion.) We thus define
$\mathcal{I}(m,n)$ to indicate descendants of $m$ that are excluded
from $n$. Thus: \begin{align*} \mathcal{I}(m,n) & =\begin{cases}
\emptyset & \text{if }n\text{ is \emph{incertae sedis}\,and }m\text{
is not}\\ \mathcal{I}(m) & \text{otherwise} \end{cases} \end{align*}

Additionally, the exclude set $\mathcal{E}(n)$ should contain the
children of all the ancestors of $n$, unless those children are
\emph{incertae sedis}. To avoid traversing all ancestors separately
for each node $n$, we note that the excluded children of any ancestors
will be in the exclude set of the parent of $n$ already. Therefore, we
can write the exclude set of a node $n$ in terms of the exclude set of
its parent:

\begin{align} \mathcal{E}(n) &
=\mathcal{E}(parent(n))\cup\left[\mathcal{I}(m,n)\big|m\in
siblings(n)\right].\label{eq:exclude-set-formula-1} \end{align} This
formula allows us to compute exclude sets via a pre-order traversal on
the taxonomy tree with exclude set for the root node defined to be
$\emptyset$.

@@

\begin{align*} \mathcal{I^{\prime}}(n) & =\begin{cases} \emptyset &
\text{if }n\in\mathcal{U}\\ \{n\} & \text{if }n\text{ is a leaf}\\{}
[\mathcal{I}^{\prime}(c)|c\in children(n)] & \text{otherwise.}
\end{cases} \end{align*} We may then define the taxa excluded from $n$
as:

@@

\bibliographystyle{upmplainnat} bibliography{otcetera}

>>>>

\documentclass[english]{article}
\usepackage[T1]{fontenc}
\usepackage[latin9]{inputenc}
\usepackage{geometry}
\geometry{verbose,tmargin=3cm,bmargin=3cm,lmargin=3cm,rmargin=3cm}
\setlength{\parskip}{6pt}
\setlength{\parindent}{0pt}
\usepackage{verbatim}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{graphicx}
\usepackage{setspace}
\linespread{1.5}
\usepackage[authoryear]{natbib}

@@

\usepackage{babel}
\usepackage{xspace}
\usepackage{color}
\usepackage[colorlinks,linkcolor=blue,citecolor=blue]{hyperref}
\begin{document}

@@

Because of the large number (over 2.6 million) of tips in the full
tree, the supertree method of \citet{redelings2017supertree} uses an
approximation approach that relies on decomposition of the full
problem into subproblems. The decomposition partitioning the list of
taxonomy splits $S(\taxonomy_{p})$ into two lists:
$\mathcal{Z}(\taxonomy_{p}),$the list of splits from the taxonomy tree
which are compatible with every rooted split in the phylogenetic
inputs, and $\mathcal{Y}(\taxonomy_{p})$, the list of taxonomic splits
that conflict with at least one phylogenetic input split. Conceptually
the approximate algorithm works by greedily adding all compatible
splits in altered ranked list of splits:
$\mathcal{S}^{\prime}=\mathcal{Z}(\taxonomy_{p})+S(T_{1})+S(T_{2})+\ldots+S(T_{n})+\mathcal{Y}(\taxonomy_{p})$.

@@

We define rooted splits for each edge by noting that removing the edge
would produce two trees. The leaf label sets of these trees the
partition the full leaf set $\mathcal{L}$ into two groups. The include
set $\mathcal{I}(e)$ which does not contain the root, and the exclude
set $\mathcal{E}(e)$ which does contain the root. Such a split may be
written \[ \mathcal{I}(e)|\bullet\mathcal{E}(e). \] where the
$\bullet$ indicates the root. The include set is the set of leaves
that are descendants of the daughter node of the edge. The exclude set
is the set of leaves that are not descendants of that node. If no taxa
are \emph{incertae sedis}, then the exclude set for a node is just the
total tip set minus the include set for the node:
$\mathcal{E}(e)  =\mathcal{L}-\mathcal{I}(e)$.

@@

We seek a semantics for supertrees with \emph{incertae sedis }taxa
that satisfy the following properties:
\begin{enumerate}
    \item An \emph{incertae sedis} node may intrude into its siblings and their
descendants.
    \item Two \emph{incertae sedis} siblings may not be
interdigitated.
    \item Any placement $T\to T^{*}$ of a tree $T$ with
\emph{incertae sedis} taxa is also accessible from a tree $T^{\prime}$
that is created from $T$ by taking a taxon, attaching it to an
ancestor, and marking it \emph{incertae sedis}. \mthcomment{MTH: I don't understand this.}
    \item The semantics of an
\emph{incertae sedis} node does not depend on assigned ranks, but only
on the tree.
    \item Semantics is based on deriving a split for each
branch of the tree.
\end{enumerate}
Here we focus on a split-based
semantics instead of a semantics based on apply of a series of edit
operations to the tree. The split-based semantics allows us to seek
trees that are simultaneously consistent with a set of split-
constraints. The use of a split-based semantics does impose some
limitations. These come primarily because in the edit-based semantics,
after we place an \emph{incertae sedis} taxon, it is no loner
\emph{incertae sedis}, whereas with the split-based semantics taxa are
either always \emph{incertae sedis} or never \emph{incertae sedis}.

@@

In order to apply supertree methods to rooted trees with
\emph{incertae sedis} taxa, we first propose a semantics of
\emph{incertae sedis }in terms of rooted splits. We define rooted
splits by noting that each edge of a tree divides the tip taxa
$\mathcal{L}$ into two groups: the include set $\mathcal{I}(e)$ which
does not contain the root, and the exclude set $\mathcal{E}(e)$ which
does contain the root. Such a split may be written \[
\mathcal{I}(e)|\bullet\mathcal{E}(e). \] where the $\bullet$ indicates
the root. If no taxa are \emph{incertae sedis}, then the exclude set
for a node is just the total tip set minus the include set for the
node: $ \mathcal{E}(n) =\mathcal{L}-\mathcal{I}(n)$.

@@

The reduced exclude set $\mathcal{E}(n)$ for node $n$ should therefore
not contain the descendants of its siblings which are marked
\emph{incertae }sedis\emph{,} but should contain the descendants of
its other siblings\emph{. }We additionally stipulate that it is not
possible for \emph{incertae sedis} taxa to intrude into their
\emph{incertae sedis} siblings. (This is not the only way to construct
a split-based semantics for \emph{incertae sedis}, but it avoids
certain complications, as noted in the discussion.) We thus define
$\mathcal{I}(m,n)$ to indicate descendants of $m$ that are excluded
from $n$. Thus:
\begin{align*}
\mathcal{I}(m,n) & =
    \begin{cases}
        \emptyset & \text{if }n\text{ is \emph{incertae sedis}\,and }m\text{
is not}\\
        \mathcal{I}(m) & \text{otherwise}
    \end{cases}
\end{align*}

Additionally, the exclude set $\mathcal{E}(n)$ should contain the
children of all the ancestors of $n$, unless those children are
\emph{incertae sedis}. To avoid traversing all ancestors separately
for each node $n$, we note that the excluded children of any ancestors
will be in the exclude set of the parent of $n$ already. Therefore, we
can write the exclude set of a node $n$ in terms of the exclude set of
its parent:
\begin{align}
    \mathcal{E}(n) & = \mathcal{E}(parent(n))\cup\left[\mathcal{I}(m,n)\big|m\in
           siblings(n)\right].\label{eq:exclude-set-formula-1}
\end{align}
This
formula allows us to compute exclude sets via a pre-order traversal on
the taxonomy tree with exclude set for the root node defined to be
$\emptyset$.

@@

\begin{align*}
    \mathcal{I^{\prime}}(n) & =
    \begin{cases}
        \emptyset & \text{if }n\in\mathcal{U}\\
        \{n\} & \text{if }n\text{ is a leaf}\\
        [\mathcal{I}^{\prime}(c)|c\in children(n)] & \text{otherwise.}
\end{cases}
\end{align*} We may then define the taxa excluded from $n$
as:

@@

\bibliographystyle{upmplainnat}
\bibliography{otcetera}
